---
title: "How to use package ATHMC: Automatically-tuned, Tempered Hamiltonian Monte Carlo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to use package ATHMC: Automatically-tuned, Tempered, Hamiltonian Monte Carlo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
sansfont: LiberationSans
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Introduction
This tutorial gives a brief introduction to the package `ATHMC` and explains how to use it using examples.
This package implements automatically-tuned, tempered Hamiltonian Monte Carlo developed by Park, J. (2025) "Sampling from high-dimensional, multimodal distribution using automatically-tuned, tempered Hamiltonian Monte Carlo" https://doi.org/10.48550/arXiv.2111.06871 

## Installation
The package can be installed from the [package author's github repository](https://github.com/joonhap/athmc) as follows.
Doing so requires having the `devtools` package installed.
```{r eval=FALSE}
install.packages('devtools') # skip this line if `devtools` is already installed.
```
Then install the `athmc` package:
```{r eval=FALSE}
devtools::install_github("joonhap/athmc")
```
The package can then be loaded as usual:
```{r}
library(ATHMC)  
```

### Example: mixture of log-polynomial distributions
We consider target densities given by
$$\pi(x) \propto e^{-\Vert x-\mu_1\Vert^\gamma} + e^{-\Vert x - \mu_2\Vert^\gamma}, \quad x\in \mathsf R^{10000},$$
where $\Vert \mu_1 - \mu_2 \Vert = 400$ and $\gamma$ is varied.
The target density and its gradient are defined as R functions as follows (`target` and `gd.target`). The function `closest_mode` outputs the closest mode (1 or 2) to the given point `x`.
```{r}
## a probability distribution with a mixture of 2 distributions whose negative log densities are polynomial in an arbitrary dimension
x.d <- 10000
modeDist <- 400 ## the distance between two modes
comp.sd <- 1 # the sd of each mixture component of the target density
set.seed(12403)
direction <- {raw.direction <- rnorm(x.d); raw.direction/sqrt(sum(raw.direction^2))}
modes <- outer(c(-1,1), modeDist*direction/2)

target <- function(x, gamma, give_log=TRUE) { ## target denstiy (up to an additive constant)
    ## gamma: polynomial degree of growth of negative target density
    if (x.d != length(x)) {
        stop("the length of x does not equal the specified dimension (x.d)") }
    lpdfs <- apply(modes, 1, function(mode) -sum((mode-x)^2)^(gamma/2))
    maxlpdf <- max(lpdfs)
    lmixpdf <- maxlpdf + log(sum(exp(lpdfs-maxlpdf))) - log(2)
    return(ifelse(give_log, lmixpdf, exp(lmixpdf)))
}

gd.target <- function(x, gamma) { ## gradient of the log of the target density
    ## gamma: polynomial degree of growth of negative target density
    if (x.d != length(x)) stop("the length of x does not equal the specificed dimension (x.d)")
    distSqs <- apply(modes, 1, function(mode) sum((mode-x)^2))
    weights <- exp(-(distSqs-min(distSqs))^(gamma/2)) # relative weights, scaled to avoid numerical instability
    return(-gamma*(weights[1]*(x-modes[1,])*distSqs[1]^(gamma/2-1)+weights[2]*(x-modes[2,])*distSqs[2]^(gamma/2-1))/sum(weights))
}

## name the mode that a point is closest to
closest_mode <- function(x) {
    dists <- apply(modes, 1, function(v) sum((x-v)^2))
    return(which.min(dists))
}
```

## Running ATHMC
```{r}
gamma <- 2 # the polynomial degree of U is 2. That is, the target is a mixture of Gaussians.
logtarget <- function(x) { target(x, gamma, TRUE) };
gradlt <- function(x) { gd.target(x, gamma) }
## Pilot run to find a mode and tune the reference leapfrog step size
set.seed(724908)
x.init <- modes[1,] + comp.sd*rnorm(x.d)
pilot_iter <- 20
pilot <- ATHMC(xinit, loglik, gd.loglik, sumstat=identity, massInv=1, niter=pilot_iter, jsize=0.01, bounds=NULL, nsteps=100, pilot_jsize_tune=TRUE, target_avg_accep=0.5, verbose=TRUE, pbar.plot=NULL)
xnewinit <- pilot$sumstat[pilot_iter+1,]
jsize <- pilot$jsize_chrono[pilot_iter+1]*.5 # reduce the tuned jsize (for robustness)
## construct a chain
niter <- 100 # length of each constructed chain
athmcOut <- ATHMC(xnewinit, logtarget, gd.logtarget, sumstat=identity, niter=niter, power=2, power.fixed=FALSE, jsize=jsize, nsteps=1000, nsteps.fixed=FALSE, maxEta=2.5, maxEta.fixed=FALSE, tune=TRUE, target_avg_accep=0.05, maxEta_tuning_method="Uthreshold", Uthreshold=Uthreshold, pbar.plot=NULL, verbose=TRUE)
```

The above code first tunes the reference leapfrog step size (denoted by $\bar{\epsilon}$ in Park (2025)) by running HMC without tempering.
This pilot run also locates a mode.

Next, using the tuned reference step size and a new initial state, tempered Hamiltonian Monte Carlo is run with adaptive tuning for
* `power`: the estimated polynomial degree of $U$ (i.e., $\hat\gamma$ in the paper),
* `maxEta`: the maximum value of the log-temperature (i.e., $\eta_*$), 
* `nsteps`: the length of the leapfrog trajectory (i.e., $K$).
The constructed Markov chain length is `niter`$+1$, including the initial state. 
The log-temperature schedule is specified as $\eta_k = \frac{2\eta_*}{K} \min(k, K-k)$, if `etaType="piecewiselinear"` (the default).
For each MCMC iteration, the parameters $\hat\gamma$, $\eta_{*}$, and $K$ are tuned, unless each parameter is excluded from tuning by setting, e.g., `power.fixed=TRUE`.
If `maxEta_tuning_method="rectangular"`, the `maxEta` parameter is tuned such that at least half of the coordinates of the simulated trajectory reach both sides of the complement of the interval `search_center[j]`$\pm$`search_scale[j]` at least once during the simulation. 
If `maxEta_tuning_method="ellipsoidal"`, then an ellipsoidal search scope is used: $\max_{k\in 0:K} \sum_{j=1}^d \{x_j\{k\} - x_j^0\}^2 / s_j^2 > d$, where $x_j\{k\}$ is the $j$-th coordinate of the $k$-th intermediate point along the simulated trajectory, $x^0$ is the `search_center`, and $s_j$ is the `search_scale`.
If `maxEta_tuning_method="Uthreshold"`, the potential function should exceed `Uthreshold` at least once along the simulated trajectory to satisfy the search criterion.

The output is a list consisting of several named entries.
```{r}
str(chain)
```

* `sumstat`: a matrix with `niter+1` rows, where each row gives the `sumstat` function applied to a state in the constructed chain.
* `jsize_chrono`: for pilot HMC run, this returns a vector of length `niter+1` recording the the reference leapfrog step size during the tuning process.
* `nsteps_chrono`: a vector of length `niter+1` recording the tuned values for the length of the eta (log-temperature) schedule at the end of each iteration. The first value gives the initial, supplied value.
* `power_chrono`: a vector of length `niter+1` recording the estimated polynomial degree of the potential function growth ($\hat\gamma$).}
* `maxEta_chrono`: a vector of length `niter+1` recording the tuned maximum value of the eta sequence.}
* `Hinc_chrono`: a vector of length `niter` recording the net increase in Hamiltonian over the trajectory constructed after tuning for that iteration.


### References
Park, J. (2025). Sampling from high-dimensional, multimodal distributions using automatically-tuned, tempered Hamiltonian Monte Carlo <https://doi.org/10.48550/arXiv.2111.06871>

